**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Scales Data: Application to the Relationship between Reality Beliefs about AI-Generated Faces and Beauty**

### 2025-05-28 paper/manuscript.qmd
changes to script

Another unknown is whether these beliefs are fully stimuli-driven or whether there are some stable individual tendencies that are at stake.

@@

In the first phase, participants viewed 109 neutral-expression photographs of faces (random order, display time of 500 ms) from the American Multiracial Face Database [@chen2021broadening]. After each image, participants rated the face on trustworthiness, familiarity, attractiveness, and beauty using visual analog scales. In the second phase, participants were informed that "about half of the previously seen images were AI-generated". The same faces were presented again in a new random order (same display time), followed by ratings of "reality" (whether they believed the image was AI-generated - left anchor - or a real photograph - right anchor). @makowski2025too showed that most stimuli yielded highly variable reality ratings, suggesting that idiosyncratic cognitive processes were at play.

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a Beta-Gate model, and the CHOCO model. For all models and each parameter, the full formula was entered: $Real\sim Sex + (1 | Participant) + (1 | Item)$ (with Sex as the main predictor and participants and items entered as random intercepts). The models were run using *brms* [@burkner2017brms] R package, and analyzed using the *easystats* collection of packages [**REFS**]. To maximize the comparability across models We used the default priors (uniform) for all models, and we ran 16 chains of 1400 iterations each on the University of Sussex High-Performance Computing (HPC) cluster.

@@

The explosion of accessibility of state-of-the-art AI tools has made it effortless to generate realistic images, including human 
faces that are often indistinguishable from real ones **[REF]**. These synthetic visuals are flooding the cyberspace across various domains, such as art, advertising and entertainment, to education and information. This technological advancement carries an important potential for misuse, such as in disinformation campaigns, scams (e.g., AI-bots, identity theft), and abuse. The democratization of such technology raises pressing concerns about the value of authenticity and the potential erosion of media trust in our increasingly *post-truth* society.

In this evolving landscape, understanding the cognitive mechanisms that underpin our judgments of reality becomes paramount. Despite the increasing prevalence of digitally altered or AI-generated content, humans still rely on certain heuristics to assess authenticity. One such heuristic might be facial attractiveness. Attractiveness appraisals are known to be automatic and unconscious [@luo2019robust; @hung2016unconscious; @hou2023neural], and carry strong real-life consequences, as demonstrated by the large body of literature on the "beauty premium" [@little2021facial; @gulati2024beautiful, **MORE**]^[As an eloquent example, @monk2021beholding reported in a large representative US sample that the magnitude of earnings disparities among white women along the perceived attractiveness continuum exceeds in magnitude the canonical black-white race gap.]

**TODO:** However, it is important to allow for the possibility of a U-shaped relationship. For instance, Peng et al. (2020) reported that both attractive and unattractive people sold more (underlining a "plainness penalty").
<!-- The Faces of Success: Beauty and Ugliness Premiums in e-Commerce Platforms -->

Although its role as a potential modulator of reality beliefs remains underexplored, Makowski et al... **TODO: detail findings of that study + other studies**.

@@

For the second sample, the procedure was relatively similar with a few key differences.
**TODO** Describe  fake Face 2. 
The main difference is that the subjective ratings were done on a X-point Likert scale, rather than analog, with a mid-point value.

>>>>

In a post-truth era [@lewandowsky2017beyond], where near-perfect simulations are increasingly feasible due to rapid technological advancements, distinguishing between what is real and what is fake poses a growing challenge [**REF**]. As these technologies advance and physical cues become indistinguishable, individuals are likely to rely more heavily on alternative sources of information -such as contextual cues and cognitive heuristic -when making reality judgments [i.e., simulation monitoring; @makowski2018cognitive, @makowski2019phenomenal]. This reliance is particularly pronounced under conditions of high ambiguity, where processing demands are higher and the atomization and decontextualization of information (especially prevalent in online environments) further complicate the discernment of authenticity.

Another unknown is whether these beliefs are fully stimuli-driven or whether there are some stable individual tendencies that are at stake. 
For instance, in the absence of clear contextual information, individual-level characteristics may increasingly guide beliefs about reality. Traits such as narcissism have been shown to influence such judgments [@makowski2025too]. Moreover, it has been proposed that the interaction between stimulus characteritics and individual differences plays a crucial role in shaping reality beliefs. For example, self-reported arousal has been found to predict the likelihood of perceiving an image as real [@azevedo2020body], while judging a stimulus as fake has been associated with emotional downregulation [@makowski2019phenomenal]. [**should probably add other examples**]

These findings underscore that what drives beliefs and judgments of reality remains unclear. Yet, understanding these mechanisms is increasingly vital given the proliferation of AI-generated content and the growing risk of misjudging the authenticity of stimuli, particularly in contexts with significant societal consequences (e.g., the influence of misinformation on democratic processes).

@@

In the first phase, participants viewed 109 neutral-expression photographs of faces (random order, display time of 500 ms) from the American Multiracial Face Database [@chen2021broadening]. After each image, participants rated the face on trustworthiness, familiarity, attractiveness, and beauty using visual analog scales. In the second phase, participants were informed that "about half of the previously seen images were AI-generated". The same faces were presented again in a new random order (same display time), followed by ratings of "reality" (whether they believed the image was fake - left anchor - or a real - right anchor). @makowski2025too showed that most stimuli yielded highly variable reality ratings, suggesting that idiosyncratic cognitive processes were at play.

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a Beta-Gate model, and the CHOCO model. For all models and each parameter, the full formula was entered: $Real\sim Sex + (1 | Participant) + (1 | Item)$ (with Sex as the main predictor and participants and items entered as random intercepts). The models were run using *brms* [@burkner2017brms] R package, and analyzed using the *easystats* collection of packages [@easystats2022]. To maximize the comparability across models We used the default priors (uniform) for all models, and we ran 16 chains of 1400 iterations each on the University of Sussex High-Performance Computing (HPC) cluster.

@@

The explosion of accessibility of state-of-the-art AI tools has made it effortless to generate realistic images, including human faces that are often indistinguishable from real ones [@miller2023ai; @nightingale2022ai; @bozkir2024can]. These synthetic visuals are flooding the cyberspace across various domains, such as art, advertising and entertainment, to education and information. This technological advancement carries an important potential for misuse, such as in disinformation campaigns, scams (e.g., AI-bots, identity theft), and abuse. The democratization of such technology raises pressing concerns about the value of authenticity and the potential erosion of media trust in our increasingly *post-truth* society.

In this evolving landscape, understanding the cognitive mechanisms that underpin our judgments of reality becomes paramount. Despite the increasing prevalence of digitally altered or AI-generated content, humans still rely on certain heuristics to assess authenticity. One such heuristic might be facial attractiveness. Attractiveness appraisals are known to be automatic and unconscious [@luo2019robust; @hung2016unconscious; @hou2023neural], and carry strong real-life consequences, as demonstrated by the large body of literature on the "beauty premium" [@little2021facial; @gulati2024beautiful;@pandey2021face; @kukkonen2024beauty]^[As an eloquent example, @monk2021beholding reported in a large representative US sample that the magnitude of earnings disparities among white women along the perceived attractiveness continuum exceeds in magnitude the canonical black-white race gap.]

**TODO:** However, it is important to allow for the possibility of a U-shaped relationship. For instance, @peng2020faces reported that both attractive and unattractive people sold more products (underlining a "plainness penalty").

Although its role as a potential modulator of reality beliefs remains underexplored, @makowski2025too found significant associations between participants’ realness ratings and facial salience. Female participants judged both highly attractive and highly unattractive faces as more likely to be real, whereas for male participants, this effect was limited to highly attractive faces. These findings offer a complementary perspective to those of @miller2023ai, who reported that AI-generated faces tend to be rated as more attractive, and that participants used attractiveness as a distinguishing cue between real and AI-generated faces. Together, these results highlight a possible bidirectional and context-dependent role of attractiveness in shaping reality judgments.

@@

For the second sample, the procedure was relatively similar with a few key differences.
Most importantly, subjective ratings were collected using a 7-point Likert scale ranging from 0 to 6, rather than a visual analog scale, which included a clear midpoint. 
Additionally, the familiarity rating was removed from the set of facial evaluations. 
In the second phase, realness judgments were also recorded on a 7-point Likert scale, this time ranging from -3 to +3, again with a defined midpoint, replacing the visual analog format used in the first study.
For this scale, the endpoints were labeled 'AI-generated' and 'Photograph', replacing the 'Fake' and 'Real' anchors used in Study 1.

There were also modifications to the instructions provided to participants. 
In Phase 1, an additional prompt was introduced, stating that the study was conducted in partnership with a young AI startup. 
This was intended to strengthen participants’ belief that the faces they were shown could plausibly be either real or AI-generated. 
Additionally, participants were explicitly asked to respond based on their immediate, first impressions—an instruction not included in the first study.

In Phase 2, participants were not informed about the proportion of AI-generated versus real faces encountered during Phase 1. 
This omission was deliberate, aimed at reducing demand characteristics and preventing participants from mechanically selecting half the faces as real and half as AI-generated. 
Instead, they were encouraged to base their responses solely on their level of confidence regarding the origin of each image.
